---
title: "FLS Unit 2"
output: html_document
---
## Locate files
```{r locate-file}
getwd()
list.files()
list.files("data")
list.files(recursive = TRUE)
```
## Setup libraries and data

```{r setup}
# Install the ggplot2 library in the console versus RStudio
library(ggplot2)
library(dplyr)

players <- read.csv("PlayersBBall.csv", header = TRUE, stringsAsFactors = FALSE)
head(players,10)
```
## Create a visual representation that summarizes the number of players in each position
```{r}
table(players$position)

players <- players %>%
  dplyr::mutate(position = trimws(as.character(position)),
                position = dplyr::na_if(position, "")) %>%
  dplyr::filter(!is.na(position))


table(players$position, useNA = "ifany")
players %>% dplyr::filter(trimws(as.character(position)) == "1") %>% dplyr::select(name, year_start, year_end, position)

players %>%
  ggplot(aes(x = position)) +
  geom_bar(fill = "blue", color = "black") +
  geom_text(
    stat = "count",
    aes(label = after_stat(count)),
    vjust = -0.3
  ) +
  labs(
    title = "Number of Basketball Players by Position",
    x = "Position",
    y = "Number of Players"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```
## Evaluate the weight and height of the various positions

```{r}
# Weight of Centers vs Forwards using boxplots

library(dplyr)
library(ggplot2)

cf <- players %>%
  filter(position %in% c("C", "F")) %>%
  mutate(weight = as.numeric(weight)) %>%
  filter(!is.na(weight))

ggplot(cf, aes(x = position, y = weight)) +
  geom_boxplot() +
  labs(
    title = "Distribution of Player Weight by Position",
    x = "Position",
    y = "Weight (lbs)"
  ) +
  theme_minimal()

library(stringr)

players <- players %>%
  mutate(
    height = as.character(height),
    feet = as.numeric(str_extract(height, "^[0-9]+")),
    inches = as.numeric(str_extract(height, "(?<=-)[0-9]+")),
    height_inches = feet * 12 + inches
  )

cf_height <- players %>%
  filter(position %in% c("C", "F")) %>%
  filter(!is.na(height_inches))

ggplot(cf_height, aes(x = position, y = height_inches)) +
  geom_boxplot() +
  labs(
    title = "Distribution of Player Height by Position",
    x = "Position",
    y = "Height (inches)"
  ) +
  theme_minimal()
```
## Use the dataframe players to evaluate additional positions
```{r}

players %>%
  filter(!is.na(height_inches)) %>%
  ggplot(aes(x = position, y = height_inches)) +
  geom_boxplot() +
  facet_wrap(~ position, scales = "free_x") +
  labs(
    title = "Distribution of Player Height by Position",
    x = "Position",
    y = "Height (inches)"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  )

```
## Use the dataframe players and additional mapping capabilities to evaluate weight and height among players
```{r}
library(dplyr)
library(ggplot2)

# Prepare data
height_weight1 <- players %>%
  mutate(
    weight = as.numeric(weight),
    height_inches = as.numeric(height_inches)
  ) %>%
  filter(!is.na(height_inches), !is.na(weight))

# Scatterplot with jitter and smooth trend line
ggplot(height_weight1, aes(x = weight, y = height_inches)) +
  geom_jitter(width = 0, height = 0.3, alpha = 0.35) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Relationship Between Player Weight and Height",
    x = "Weight (lbs)",
    y = "Height (inches)"
  ) +
  theme_minimal()

```
## Use the dataframe players and additional mapping capabilities to evaluate weight and height across positions
```{r}
library(tidyr)

height_weight2 <- players %>%
  mutate(
    weight = as.numeric(weight),
    height_inches = as.numeric(height_inches)  
  ) %>%
  filter(!is.na(height_inches), !is.na(weight)) %>%
  filter(position != "")

ggplot(height_weight2, aes(x = height_inches, y = weight)) +
  geom_jitter(width = 0.3, height = 0, alpha = 0.35) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~ position) +
  labs(
    title = "Height vs Weight by Position",
    x = "Height (inches)",
    y = "Weight (lbs)"
  ) +
  theme_minimal()

```
## Examine the claims of a historian
```{r}
library(dplyr)
library(ggplot2)

players <- players %>%
  mutate(
    decade = floor(year_start / 10) * 10
  )

ggplot(players, aes(x = year_start, y = height_inches, color = factor(decade))) +
  geom_point(alpha = 0.35) +
  geom_smooth(se = FALSE, color = "black") +
  labs(
    title = "Player Height Over Time, Colored by Decade",
    x = "Year",
    y = "Height (inches)",
    color = "Decade"
  ) +
  theme_minimal()

```
## Create a 3D plot
```{r}
library(dplyr)

height_weight_3d <- players %>%
  mutate(
    year_start = as.numeric(year_start),
    weight = as.numeric(weight),
    height_inches = as.numeric(height_inches),
    position = as.factor(position)
  ) %>%
  filter(!is.na(year_start), !is.na(weight), !is.na(height_inches)) %>%
  filter(position != "")

library(plotly)

plot_ly(
  data = height_weight_3d,
  x = ~year_start,
  y = ~weight,
  z = ~height_inches,
  color = ~position,
  type = "scatter3d",
  mode = "markers",
  marker = list(size = 3, opacity = 0.6)
) %>%
  layout(
    title = "3D Scatter: Year vs Weight vs Height (Colored by Position)",
    scene = list(
      xaxis = list(title = "Year"),
      yaxis = list(title = "Weight (lbs)"),
      zaxis = list(title = "Height (inches)")
    )
  )
```
## Chosing the violin plot + jitter to provide additional insight on height and weight
```{r}
library(ggplot2)
library(dplyr)

players %>%
  filter(position != "", !is.na(height_inches)) %>%
  ggplot(aes(x = position, y = height_inches, fill = position)) +
  geom_violin(alpha = 0.6, trim = FALSE) +
  geom_jitter(width = 0.15, alpha = 0.25, size = 0.7) +
  labs(
    title = "Height Distribution by Position (Violin + Points)",
    x = "Position",
    y = "Height (inches)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

```
## Create a visualization for the dataset EducationIncome
```{r}
library(ggplot2)
library(dplyr)

getwd()
list.files()

EducationIncome <- read.csv("Education_Income.csv", header = TRUE, stringsAsFactors = FALSE)
head(EducationIncome,10)

# Create a categorical ordering of the data since there are 3 different types of text values

EducationIncome <- EducationIncome %>%
  mutate(
    Educ_clean = case_when(
      Educ == "12" ~ "12",
      Educ == "13-15" ~ "13–15",
      Educ == "16" ~ "16",
      Educ == ">16" ~ ">16",
      TRUE ~ NA_character_
    ),
    Educ_clean = factor(
      Educ_clean,
      levels = c("12", "13–15", "16", ">16"),
      ordered = TRUE
    )
  )
EducationIncome %>%
  group_by(Educ_clean) %>%
  summarise(
    n = n(),
    mean_income = mean(Income2005, na.rm = TRUE),
    median_income = median(Income2005, na.rm = TRUE)
  )
library(ggplot2)
library(scales)

ggplot(EducationIncome, aes(x = Educ_clean, y = Income2005)) +
  geom_boxplot() +
  scale_y_continuous(labels = dollar_format()) +
  labs(
    title = "Income Distribution by Education Level",
    x = "Education Level",
    y = "Income (2005)"
  ) +
  theme_minimal()



```
